High-Performance Primitives for Distributed Memory Architectures using Asynchronous Events

### 2012-11-13 doc/llr-pldi2013/impl.tex
update
Lots of small edits

%Due to space constraints, we will not cover every part of our implementation in
%detail, instead limiting our discussion to the three most interesting aspects:
%events, locks, and reduction instances.
Due to space constraints we cover only the most interesting parts of the implementation in detail.
We limit our discussion to three aspects:
events, locks, and reduction instances.

@@

In addition to normal physical region instances, our implementation supports a special reduction-only
instance that differ in two important ways.  First, the per-element storage in a reduction-only instance
is sized to hold the ``right-hand side'' of the reduction operation (e.g. the $v$ in an operation like
$struct.field\ \text{+=}\ v$), allowing a more compact representation.  Second, individual reductions are applied
(atomically) to the local instance, which can then be sent as a large batched reduction to the target
instance.  When multiple reductions are made to the same element, they can be {\em folded} together locally,
further reducing the inter-node communication needed.  Note that the folding operation is not always
identical to the reduction operation itself.  For example, if the reduction operation is exponentiation,
the corresponding fold operation is multiplication, according to the following identity:
$$(r[i]\ \text{**=}\ a)\ \text{**=}\ b \quad \Leftrightarrow \quad r[i]\ \text{**=}\ (a * b).$$

>>>>

%Due to space constraints, we will not cover every part of our implementation in
%detail, instead limiting our discussion to the three most interesting aspects:
%events, locks, and reduction instances.
%Due to space constraints we cover only the most interesting parts of the implementation in detail.
%We limit our discussion to three aspects:
%events, locks, and reduction instances.
Due to space constraints, we only describe the implementation of features
that are technical contributions.  We cover in detail
the implementation of events, deferred locks, and reduction instances.

@@

In addition to normal physical region instances, our implementation supports a special reduction-only
instance that differs in two important ways.  First, the per-element storage in a reduction-only instance
is sized to hold the ``right-hand side'' of the reduction operation (e.g. the $v$ in an operation like
$struct.field\ \text{+=}\ v$), allowing a more compact representation.  Second, individual reductions are applied
(atomically) to the local instance, which can then be sent as a large batched reduction to the target
instance.  When multiple reductions are made to the same element, they can be {\em folded} together locally,
further reducing the inter-node communication needed.  Note that the folding operation is not always
identical to the reduction operation itself.  For example, if the reduction operation is exponentiation,
the corresponding fold operation is multiplication, according to the following identity:
$$(r[i]\ \text{**=}\ a)\ \text{**=}\ b \quad \Leftrightarrow \quad r[i]\ \text{**=}\ (a * b).$$

### 2012-11-13 doc/llr-pldi2013/interface.tex
Lots of small edits

Events and barriers express ordering properties between operations, but 
%in many cases 
ordering is often too strict a requirement.  For many applications access to data need only be atomic and
not necessarily ordered.  Traditionally locks have been used to serialize access to
data.  However, all lock implementations that we are aware of require either blocking
or spinning, neither of which composes well with asynchronous operations.
{\em Deferred locks} are a new synchronization mechanism that allows for synchronization
without blocking or spinning.
%in an completely non-blocking environment.

@@

If the only operations a task will perform on a instance are reductions, a special reduction-only
instance can be created (line 55).  Rather than applying each reduction separately to the target instance,
the reduction operations can be accumulated in the reduction instance.  Using the {\tt reduce\_to} 
method (line 64), these accumulated reductions can be applied in bulk to another instance, which may be either
a normal instance or another reduction-only instance.  The use of reduction instances can result 
in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
Furthermore, reduction-only instances are often smaller than normal instances, potentially improving
cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device
memory).  For example, if reductions are being made to a single field in a larger structure, the 
reduction-only instance only needs to store that field and not the other fields that won't be 
modified by the reduction operation.

>>>>

Events and barriers express ordering properties between operations, but 
%in many cases 
ordering is often too strict a requirement.  For many applications access to data need only be atomic and
not necessarily ordered.  Traditionally locks have been used to serialize access to
data.  However, lock implementations in existing parallel programming models require either blocking
or spinning, neither of which composes well with asynchronous operations.
{\em Deferred locks} are a new synchronization mechanism that allows for synchronization
without blocking or spinning.
%in an completely non-blocking environment.

@@

If the only operation a task will perform on an instance is a reduction, a special reduction-only
instance can be created (line 55).  Rather than applying each reduction separately to the target instance,
the reduction operations can be accumulated in the reduction instance.  Using the {\tt reduce\_to} 
method (line 64), these accumulated reductions can be applied in bulk to another instance, which may be either
a normal instance or another reduction-only instance.  The use of reduction instances can result 
in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
Furthermore, reduction-only instances are often smaller than normal instances, potentially improving
cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device
memory).  For example, if reductions are being made to a single field in a larger structure, the 
reduction-only instance only needs to store that field and not the other fields that won't be 
modified by the reduction operation.

### 2012-11-13 doc/llr-pldi2013/intro.tex
Lots of small edits

We propose a new low-level interface for programming modern
supercomputers that is fully asynchronous. In our system, one can
asynchronously compose computation, data movement, and
synchronization.  The basic mechanism that enables composition of 
asynchronous operations
is an {\em event} primitive.  Events provide a mechanism
for naming a point in the future when an asynchronous operation
(computation, data copy, or synchronization) will complete.  Every
call to perform an operation $p$ in our runtime returns immediately
with an event that {\em triggers} when $p$ has completed.
Furthermore, every call to perform an operation $p$ can take as a
precondition an event that must trigger before $p$ can begin.
%Any number of operations can be depedendent on a single event, and
%a events can be merged to allow an operation to wait for multiple
%events before it begins.
Events can be merged into new events and multiple operations
can use an event as a precondition.
  Using
events clients can compose arbitrary chains of dependent
operations, which the runtime is free to execute in any way that
respects the event dependences.  With this freedom the runtime can
schedule operations to optimize throughput while hiding long
latency operations.

@@

\item Our approach depends on the efficient handling of very large numbers of events.
We describe a distributed implementation of our interface that requires no global coordination
and has modest local storage costs.  We describe an optimized implementation of events, 
deferred locks, and reductions (Section~\ref{sec:impl}).

>>>>

We propose a new low-level interface for programming modern
supercomputers that is fully asynchronous. In our system, one can
asynchronously compose computation, data movement, and
synchronization.  The basic mechanism that enables composition of 
asynchronous operations
is an {\em event} primitive.  Events provide a mechanism
for naming a point in the future when an asynchronous operation
(computation, data copy, or synchronization) will complete.  Every
call to perform an operation $p$ in our runtime returns immediately
with an event that {\em triggers} when $p$ has completed.
Furthermore, every call to perform an operation $p$ can take as a
precondition an event that must trigger before $p$ can begin.
%Any number of operations can be depedendent on a single event, and
%a events can be merged to allow an operation to wait for multiple
%events before it begins.
Events can be merged into new events and multiple operations
can use an event as a precondition.
Using events, clients can compose arbitrary chains of dependent
operations, which the runtime is free to execute in any way that
respects the event dependences.  With this freedom the runtime can
schedule operations to optimize throughput while hiding long
latency operations.

@@

\item We describe an implmentation of events based on a {\em generational event}
data structure that compactly represents a large number of events.  Generational
events also amortize the cost of inter-node communication when multiple operations are dependent
on the same event (Section~\ref{sec:impl}).
%%\item Our approach depends on the efficient handling of very large numbers of events.
%%We describe a distributed implementation of our interface that requires no global coordination
%%and has modest local storage costs.  We describe an optimized implementation of events, 
%%deferred locks, and reductions (Section~\ref{sec:impl}).
